### 2015-08-04 background.tex
pass over the paper addressing Zvonimir's feedback

Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program. Initially, $\mathit{CLS}_T$ is empty for every thread $T$, because the threads do not hold any locks on program start. The lockset $\mathit{LS}_s$ for each variable $s$ is initialized to the set of all locks manipulated by the program, because initially each access to $s$ has (vacuously) been protected by every lock. The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
%
After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
%
When $T$ accesses shared variable $s$, any locks that are not common to $\mathit{LS}_s$ and $\mathit{CLS}_T$ are removed from $\mathit{LS}_s$.
If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.

>>>>

Lockset analysis for a concurrent program starts by creating a \emph{current} lockset $\mathit{CLS}_T$ for each thread $T$ of the program, and a lockset $\mathit{LS}_s$ for each shared variable $s$ used in the program. Initially, $\mathit{CLS}_T$ is empty for every $T$, because the threads do not hold any locks on program start. The lockset $\mathit{LS}_s$ for each variable $s$ is initialized to the set of all locks manipulated by the program, because initially each access to $s$ has (vacuously) been protected by every lock. The program is executed as usual (with threads scheduled according to the OS scheduler), except that instrumentation is added to update locksets as follows.
%
After each \emph{lock} and \emph{unlock} operation by $T$, $\mathit{CLS}_T$ is updated to reflect the locks currently held by $T$.
%
When $T$ accesses a shared variable $s$, $\mathit{LS}_s$ is updated to the intersection of $\mathit{LS}_s$ with $\mathit{CLS}_T$, which removes any locks that are not common to the two locksets.
%
If $\mathit{LS}_s$ becomes empty as a result, a warning is issued that the access to $s$ may be unprotected.

### 2015-08-04 corral.tex
pass over the paper addressing Zvonimir's feedback

We have implemented two different \texttt{yield} instrumentations in \whoop: \yieldcoarse and \yieldmr.
%
The first instruments \texttt{yield} statements in a binary fashion: if \whoop proves an entry point pair as race free, then it will instrument a \texttt{yield} only after each lock and unlock statement of the pair; else if \whoop finds that a pair might race, then it will instrument a \texttt{yield} after all visible operations of the pair.
%
\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region that might race in the pair (regardless if the pair has not been fully proven as race-free), and after each lock and unlock. In our experiments (see \S\ref{evaluation}), \yieldmr significantly outperforms \yieldcoarse.

Our partial-order reduction can be used to accelerate \corral for arbitrary bug-finding in concurrent programs. In the remainder of this paper, though, we only regard races as bugs and use \corral solely to look for races.

>>>>

We have implemented two different \texttt{yield} instrumentations in \whoop: \yieldcoarse and \yieldmr.
%
The first instruments \texttt{yield} statements in a binary fashion: if \whoop proves an entry point pair as race free, then it will instrument a \texttt{yield} only after each lock and unlock statement of the pair; else if \whoop finds that a pair might race, then it will instrument a \texttt{yield} after all visible operations of the pair.
%
\yieldmr is a finer-grained instrumentation: it instruments a \texttt{yield} only after each access to a memory region that might race in the pair (regardless if the pair has not been fully proven as race-free), and after each lock and unlock. In our experiments (see \S\ref{evaluation}), \yieldmr is significantly faster than \yieldcoarse.

Our partial-order reduction is able in principle to accelerate \corral for \emph{arbitrary} bug-finding in concurrent programs. Although we did preliminary explorations in this direction, identifying useful safety properties to check was challenging since there are typically no assertions in drivers. Thus, in the remainder of this paper, we only regard data races as bugs and use \corral solely to look for data races.

### 2015-08-04 evaluation.tex
pass over the paper addressing Zvonimir's feedback

\noindent\textbf{Race-Checking }
%
Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found). Using a higher csb did not uncover any further races; this might mean that races in our benchmarks require only one or two context-switches to manifest, or that \corral hit its bounds before discovering a deeper bug. \corral did not discover any races that \whoop did not already report.  This is expected, as \whoop aims for soundness, and increases our confidence in the implementation of \whoop.

We can see that \whoop reports more races than \corral does. This is because \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false positives.  On the other hand, \corral is precise, but can miss races because only a limited number of context-switches are considered.  Another issue with \corral is loop coverage due to unsound loop unrolling. To tackle this, we enable the built-in loop over-approximation described in~\cite{lal2014powering}. This can potentially lead \corral to report false bugs, but we have not seen this in practice. Finally, when we apply \corral to a pair of entry points, we just check the specific pair and do not account for the effects of other threads (see \S\ref{corral}); this can also cause \corral to miss some races.

>>>>

\noindent\textbf{Race-Checking }
%
Table~\ref{tab:stats} presents statistics for our benchmarks: lines of code (LoC); number of possible entry point pairs (\#Pairs); number of SMACK memory regions (\#MRs); number of racy pairs (\#Racy Pairs) and number of racy memory regions (\#Racy MRs) reported by \whoop; and number of data races discovered by \corral using a context-switch bound (csb) of 2 (\#Races Found).\footnote{The number of racy memory regions can be less than the number of races found by \corral, because \whoop might find that a memory region is racy, but the same memory region might race in many points in the program.} Using a higher csb did not uncover any further races; this might mean that all races in our benchmarks can manifest with a csb of $\leq$ 2, or that \corral hit its recursion depth bound before discovering a deeper bug.

We can see that \whoop reports more races than \corral does. This is because \whoop employs an over-approximating shared state abstraction to conservatively model the effects of additional threads when analyzing an entry point pair, and because lockset analysis is inherently imprecise; both factors can lead to false positives.  On the other hand, \corral is precise, but can miss races because only a limited number of context-switches are considered.  Another issue with \corral is loop coverage due to unsound loop unrolling. To tackle this, we enable the built-in loop over-approximation described in previous work~\cite{lal2014powering}. This can potentially lead \corral to report false bugs, but we have not seen this in practice. Furthermore, when we apply \corral to a pair of entry points, we just check the specific pair and do not account for the effects of other threads (see \S\ref{corral}); this can also cause \corral to miss some races. Standalone \corral did not discover any data races that \whoop did not already report.  This is expected, as \whoop aims for soundness, and increases our confidence in the implementation of \whoop.

### 2015-08-04 intro.tex
pass over the paper addressing Zvonimir's feedback

Concurrency bugs are exacerbated by the complex and hostile environment in which drivers operate~\cite{corbet2005linux}. The OS can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may remove a device (hot-plugging) while some operation is still running.  These scenarios can cause \emph{data races} if insufficient synchronization mechanisms are in place to control concurrent access to shared resources.
%
Data races are a source of undefined behavior in C~\cite[p.\ 38]{iso/iec11}, and lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and fix, especially in the context of complex operating systems.
%
Several techniques have been successfully used to analyze device drivers~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral}, but most focus on generic sequential program properties and protocol bugs. Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races. Techniques that can detect races in drivers~\cite{dawson2003racerx, qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability. Thus, there is a clear need for new tools that are able to detect races efficiently and precisely.

>>>>

Concurrency bugs are exacerbated by the complex and hostile environment in which drivers operate~\cite{corbet2005linux}. The OS can invoke drivers from multiple threads, a hardware device can issue interrupt requests that cause running processes to block and switch execution context, and the user may remove a device (hot-plugging) while some operation is still running.  These scenarios can cause \emph{data races} if insufficient synchronization mechanisms are in place to control concurrent access to shared resources.
%
Data races are a source of undefined behavior in C~\cite[p.\ 38]{iso/iec11}, and lead to nondeterministically occurring bugs that can be hard to reproduce, isolate and fix, especially in the context of complex operating systems.
%
Several techniques have been successfully used to analyze device drivers~\cite{ball2006thorough, clarke2004predicate, engler2000checking, henzinger2002temporal, cook2006termination, kuznetsov2010testing, renzelmann2012symdrive, lal2012corral}, but most focus on generic sequential program properties and protocol bugs. Linux kernel analyzers, such as sparse~\cite{corbet2004sparse}, coccinelle~\cite{padioleau2008doc} and lockdep~\cite{corbet2006lock}, can find deadlocks in kernel source code, but are unable to detect races. Techniques that can detect races in drivers~\cite{dawson2003racerx, qadeer2004kiss, pratikakis2006locksmith, voung2007relay, lal2012corral} are usually either overly \emph{unsound} (i.e.\ can miss real bugs) or \emph{imprecise} (i.e.\ can report false bugs), and typically sacrifice precision for scalability. Thus, there is a clear need for new tools that can detect races efficiently and precisely.

### 2015-08-04 whoop.tex
pass over the paper addressing Zvonimir's feedback

Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.  For each such pair, we use symbolic verification to check whether it is possible for the pair to race on a shared memory location. We soundly model the effects of additional threads by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.  Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.
\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself. However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel (for performance), although we leave this for future work.}
The trade-off is that a quadratic number of entry point pairs must be checked.  In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

@@

The simple concurrent programming model of \S\ref{sec:symbolicpairwise} is deliberately idealistic to make it easy to describe our symbolic verification technique. In practice, Linux drivers are written in C, our technique does not know up-front which are the driver entry points, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers. We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop\footnote{\url{https://github.com/pdeligia/whoop}}, a practical, fully automatic tool for detecting data races in drivers.

\noindent\textbf{Architecture }
%
Figure~\ref{fig:whoop} depicts the \whoop toolchain. The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races. Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur\footnote{\url{https://github.com/mc-imperial/chauffeur}}, Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral). Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and \S\ref{corral}).

@@

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique constant Boogie variable. The reason behind this transformation is that representing all locks statically, instead of their original SMACK pointer-based representation, helps \whoop perform various internal instrumentations and optimizations.
%
Currently, \whoop only supports mutexes and spinlocks that are available in the Linux kernel APIs. If \whoop cannot infer a lock, e.g. because it was created dynamically or was indexed from an array of locks, it will exit with a warning. There are two reasons behind this: (i) it is arguably hard to detect such locks via static analysis; and (ii) a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.

>>>>

Our approach considers, for a given driver, every pair of entry points that can potentially execute concurrently.  For each such pair, we use symbolic verification to check whether it is possible for the pair to race on a shared memory location. We soundly model the effects of additional threads by treating the driver shared state abstractly: when an entry point reads from the shared state, a nondeterministic value is returned.  Restricting to pairs of entry points, rather than analyzing all entry points simultaneously, exploits the fact that data races occur between pairs of threads and limits the complexity of the generated verification conditions.\footnote{In principle, our approach could be applied at a coarser level of granularity: e.g.\ by considering all entry points one after the other, taking into account that an entry point can race with itself. However, using pairwise analysis has the additional advantage that it enables us to easily run the analysis for each pair in parallel (for performance), although we leave this for future work.}
%
The trade-off is that a quadratic number of entry point pairs must be checked.  In \S\ref{sec:implementation} we discuss optimizations based on device driver domain knowledge to reduce the number of pairs to some extent.

@@

The simple concurrent programming model of \S\ref{sec:symbolicpairwise} is deliberately idealistic to make it easy to describe our symbolic verification technique. In practice, Linux drivers are written in C, our technique does not know up-front which are the driver entry points, drivers do not work with a cleanly specified set of named locks, and rather than having a given set of named shared variables, we have arbitrary memory accesses via pointers. We now explain how we have taken the conceptual ideas from \S\ref{sec:symbolicpairwise} and used them to build \whoop,\footnote{\url{https://github.com/pdeligia/whoop}} a practical, fully automatic tool for detecting data races in drivers.

\noindent\textbf{Architecture }
%
Figure~\ref{fig:whoop} depicts the \whoop toolchain. The input to \whoop is a Linux driver written in C, together with an environmental model\footnote{This consists of stub header files modeling relevant Linux kernel APIs.} that is required to ``close'' the driver so that it can be analyzed for races. Initially, \whoop uses three LLVM\footnote{\url{http://llvm.org}}-based tools, Chauffeur,\footnote{\url{https://github.com/mc-imperial/chauffeur}} Clang\footnote{\url{http://clang.llvm.org}} and SMACK\footnote{\url{https://github.com/smackers/smack}}~\cite{rakamaric2014smack}, to translate the driver and its environmental model into an abstract program written in the Boogie intermediate verification language (IVL)~\cite{deline2005boogiepl}, a simple imperative language with well-defined semantics that is used as the input to a number of cutting-edge verifiers (e.g.\ Boogie and \corral). Next, \whoop instruments and sequentializes the program to perform symbolic pairwise lockset analysis (see Figure~\ref{fig:whoop} -- B and \S\ref{sec:symbolicpairwise}) using the Boogie verification engine. After the verification phase ends, \whoop can exploit any inferred race-freedom guarantees to accelerate precise race-checking with \corral (see Figure~\ref{fig:whoop} -- C and \S\ref{corral}).

@@

\noindent\textbf{Identifying Locks }
%
When the instrumentation phase begins, \whoop performs an inter-procedural static analysis (at Boogie IVL level) to identify all available locks and rewrite each one (both at declaration and at all access sites) to a unique Boogie constant. Representing all locks statically, instead of their original SMACK pointer-based representation, makes it easier for \whoop to perform lockset instrumentation and lockset-based invariant generation.
%
Currently, \whoop supports Linux kernel mutex and spinlock operations.  In addition, it is easy to enhance it with knowledge of other locking primitives.
%
If \whoop cannot reliably infer a unique lock instance (e.g., because it is stored in an unbounded data structure of locks such as an array or a list), it will exit with a warning. However, we have never encountered this in practice since a small, fixed number of locks is advocated by Linux experts as good practice when developing drivers~\cite[p.\ 123]{corbet2005linux}.
